---
title: "9.1. tissue vs urine re-comparison: ES and MSigDB"
author: "YD-X"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
    number_sections: true
    latex_engine: xelatex
  html_document: 
    toc: true
    number_sections: true
---

This script is to process the data using the best method (*i.e.*, the method with the highest $I_{TUC}$) and compare the gene set enrichment between tissue and urine samples. The gene sets include epithelial and stromal gene sets that we curated and the gene sets retrieved from The Molecular Signature Database (MSigDB).

# Preparation

```{r message=FALSE, warning=FALSE}
library(magrittr)
library(dplyr)
library(ggplot2)
library(stringr)
library(reshape2)
library(cowplot)
library(ggpubr)
library(GSVA)
library(pheatmap)
library(coin)
library(rstatix)
library(openxlsx)
roxygen2::roxygenize("../8. TUCI re-computation/functions for TUCI")
```

Packages not always used

```{r}
# library(sva)
# library(msigdbr)
# library(clusterProfiler)
# library(enrichplot)
# library(RCy3)
# library(biomaRt)
```

```{r}
setwd("E:/study/Queens/OneDrive - Queen's University/project codes and results/9. tissue vs urine re-comparison")
```

```{r message=FALSE}
load("9.1. tissue vs urine re-comparison--ES and MSigDB.RData")
```

# Load data

```{r}
load("../4. PRONTO ES gene expression & selection in GEO sets/PRONTO ES features in GEO sets.RData")

load("../6. PRONTO data & matched urine tissue data normalization/Limited paired sample processing.RData")

load("../8. TUCI re-computation/8.2.3 TUCI of genes test.RData")

rm(list = ls(all.names = T)[
  !ls(all.names = T) %in% 
    c("PRONTO.NGSb1.samplerow","sample.info",
      "gene_ids","PRONTO_epi.final","PRONTO_str.final")])

gene.id <- gene_ids
epi.gene.set <- PRONTO_epi.final
str.gene.set <- PRONTO_str.final
exp.ori.sample_row <- PRONTO.NGSb1.samplerow

rm(gene_ids,PRONTO_epi.final,
   PRONTO_str.final,
   PRONTO.NGSb1.samplerow)
```

# Data pre-processing

Batch effect correction of expression data

```{r}
# Remove samples with NAs in the GS information.
.GS_na.sample <- 
  sample.info %>%
  dplyr::filter(is.na(GS)) %>% 
  .$SampleID.readable

sample.info %<>%
  dplyr::filter(!SampleID.readable %in% .GS_na.sample)

exp.ori.sample_row %<>%
  dplyr::filter(!rownames(.) %in% .GS_na.sample)

# Required inputs: batch and covariates
.combat.sample.batch <- 
  exp.ori.sample_row %>% 
  rownames() %>% 
  str_extract("(?<=Q-\\d{3}\\s)(T|U)")

.combat.sample.mod <- 
  model.matrix(data = sample.info,
               ~as.factor(GS))

# Correct batch effect
exp.proc.sample_col <- 
  ComBat(dat = exp.ori.sample_row %>% t,
         batch = .combat.sample.batch,
         mod = .combat.sample.mod,
         mean.only = T,par.prior = F) %>% 
  data.frame(check.names = F) %>% 
  dplyr::arrange(rownames(.))

exp.proc.sample_row <- 
  exp.proc.sample_col %>% 
  t %>% 
  data.frame(check.names = F)
```

Extract matched samples

```{r}
mtu.sample <- 
  sample.info %>% 
  dplyr::filter(match) %>% 
  dplyr::select(contains(
    c("PatientID","SampleID.readable","SampleType")))

exp.proc.mtu <- 
  exp.proc.sample_col %>% 
  .[,colnames(.) %in% mtu.sample$SampleID.readable]
```

```{r}
mtu.sample %>%
  dplyr::filter(SampleType == "urine") %>% 
  nrow
```

```{r}
mtu.sample %>%
  dplyr::filter(SampleType == "tissue") %>% 
  nrow
```

```{r}
mtu.sample$PatientID %>%
  unique() %>% 
  length()
```

There are 40 matched tissue samples and 46 matched urine samples from 39 patients.

# Method validation

Validate the final processing method. Calculate rho values between the mtu samples and p values of differential genes.

```{r}
# Calculate Rho and p values
tuci.compo.ori <- 
  calc_indiv_sample_tuci_compo(
    exp = exp.ori.sample_row,
    sample.info = sample.info)

tuci.compo.proc <- 
  calc_indiv_sample_tuci_compo(
    exp = exp.proc.sample_row,
    sample.info = sample.info)

# Organize results

## Rho
tuci.rho.change <- 
  data.frame(
    original = tuci.compo.ori$mtu.rho$rho,
    processed = tuci.compo.proc$mtu.rho$rho,
    patient.id = rownames(tuci.compo.ori$mtu.rho)) %>% 
  reshape2::melt(
    id.vars = "patient.id",
    variable.name = "dataset",
    value.name = "value")

## tissue p values
tuci.tissue_p.change <- 
  data.frame(
    gene = tuci.compo.ori$tissue.deg.p$gene,
    original = tuci.compo.ori$tissue.deg.p$p,
    processed = tuci.compo.proc$tissue.deg.p$p) %>% 
  dplyr::mutate(
    changed = case_when(
      original == processed ~ "unchanged",
      original != processed ~ "changed")) %>% 
  reshape2::melt(
    id.vars = c("gene","changed"),
    variable.name = "dataset",
    value.name = "value")

## Significant genes in tissue
sig_genes.tissue <- 
  tuci.compo.ori$tissue.deg.p %>% 
  dplyr::filter(p < 0.05) %>% 
  .$gene %>% 
  as.character()

tuci.tissue_p.change.sig <- 
  tuci.tissue_p.change %>% 
  dplyr::filter(gene %in% sig_genes.tissue) %>% 
  reshape2::dcast(
    formula = gene + changed ~ paste0("p.",dataset)) %>% 
  dplyr::arrange(p.original)

## urine p values
tuci.urine_p.change <- 
  data.frame(
    gene = tuci.compo.ori$urine.deg.p$gene,
    original = tuci.compo.ori$urine.deg.p$p,
    processed = tuci.compo.proc$urine.deg.p$p) %>% 
  dplyr::mutate(
    changed = case_when(
      original == processed ~ "unchanged",
      original != processed ~ "changed")) %>% 
  reshape2::melt(
    id.vars = c("gene","changed"),
    variable.name = "dataset",
    value.name = "value")

## Significant genes in urine
sig_genes.urine <- 
  tuci.compo.ori$urine.deg.p %>% 
  dplyr::filter(p < 0.05) %>% 
  .$gene %>% 
  as.character()

tuci.urine_p.change.sig <- 
  tuci.urine_p.change %>% 
  dplyr::filter(gene %in% sig_genes.urine) %>% 
  reshape2::dcast(
    formula = gene + changed ~ paste0("p.",dataset)) %>% 
  dplyr::arrange(p.original)
```

## Organize significant gene results

```{r}
write.xlsx(
  list(tuci.tissue_p.change.sig,
       tuci.urine_p.change.sig),
  "results/validation-p of significant genes.xlsx",
  sheetName = c("tissue","urine"))
```

## Visualization

Rho

```{r}
ggplot(
  data = tuci.rho.change,
  mapping = aes(
    x = dataset, y = value, fill = dataset)) +
  stat_summary(
    fun.min = function(x){quantile(x)["75%"]+1.5*sd(x)},
    fun.max = function(x){quantile(x)["25%"]-1.5*sd(x)},
    geom = "errorbar",
    position = position_dodge(width = 0.8),
    width = 0.4,color = "black",linewidth = 0.5,) +
  geom_boxplot(
    position = position_dodge(0.8)) +
  geom_point(
    position = position_dodge(0.8)) +
  geom_line(
    mapping = aes(
      group = patient.id)) +
  scale_fill_manual(
    labels = c("original data",
               "processed data"),
    values = c("grey50","chocolate")) +
  xlab(NULL) +
  ylab(NULL) +
  ggtitle("Correlation Coefficient Rho\nin MTU samples") +
  scale_x_discrete(
    labels = c("original\ndata",
               "processed\ndata")) +
  theme(
    axis.line = element_line(),
    panel.background = element_blank(),
    plot.title = element_text(
      size = 20,face = "bold",hjust = 0),
    axis.text.x = element_text(
      size = 20,hjust = 0.5),
    axis.text.y = element_text(
      size = 20),
    legend.title = element_blank(),
    legend.position = "none",
    legend.key = element_rect(
      fill = "white",color = "white"),
    legend.key.height = unit(1,"cm"),
    legend.key.width = unit(0.5,"cm"),
    legend.text = element_text(size = 20))
```

tissue p values

```{r}
ggplot(
  data = tuci.tissue_p.change,
  mapping = aes(
    x = dataset, y = value, fill = dataset)) +
  stat_summary(
    fun.min = function(x){quantile(x)["75%"]+1.5*sd(x)},
    fun.max = function(x){quantile(x)["25%"]-1.5*sd(x)},
    geom = "errorbar",
    position = position_dodge(width = 0.8),
    width = 0.4,color = "black",linewidth = 0.5,) +
  geom_boxplot(
    position = position_dodge(0.8)) +
  geom_point(
    position = position_dodge(0.8)) +
  scale_fill_manual(
    labels = c("original data",
               "processed data"),
    values = c("grey50","chocolate")) +
  geom_line(
    mapping = aes(
      group = gene,
      color = changed)) +
  scale_color_manual(
    labels = c(
      paste0("changed\n(",
             tuci.tissue_p.change$changed %>% 
               table %>% 
               .["changed"]/2,
             " genes)"),
      paste0("unchanged\n(",
             tuci.tissue_p.change$changed %>% 
               table %>% 
               .["unchanged"]/2,
             " genes)")),
    values = c("lightpink1","khaki")) +  
  xlab(NULL) +
  ylab(NULL) +
  guides(fill = "none") +
  ggtitle("p-values of differential expression\nin tissue samples") +
  scale_x_discrete(
    labels = c("original\ndata",
               "processed\ndata")) +
  theme(
    axis.line = element_line(),
    panel.background = element_blank(),
    plot.title = element_text(
      size = 20,face = "bold",hjust = 0),
    axis.text.x = element_text(
      size = 20,hjust = 0.5),
    axis.text.y = element_text(
      size = 20),
    legend.title = element_blank(),
    legend.key = element_rect(
      fill = "white",color = "white"),
    legend.key.height = unit(3,"cm"),
    legend.key.width = unit(0.5,"cm"),
    legend.text = element_text(size = 20))
```

urine p-values

```{r}
ggplot(
  data = tuci.urine_p.change,
  mapping = aes(
    x = dataset, y = value, fill = dataset)) +
  stat_summary(
    fun.min = function(x){quantile(x)["75%"]+1.5*sd(x)},
    fun.max = function(x){quantile(x)["25%"]-1.5*sd(x)},
    geom = "errorbar",
    position = position_dodge(width = 0.8),
    width = 0.4,color = "black",linewidth = 0.5,) +
  geom_boxplot(
    position = position_dodge(0.8)) +
  geom_point(
    position = position_dodge(0.8)) +
  scale_fill_manual(
    labels = c("original data",
               "processed data"),
    values = c("grey50","chocolate")) +
  geom_line(
    mapping = aes(
      group = gene,
      color = changed)) +
  scale_color_manual(
    labels = c(
      paste0("changed\n(",
             tuci.urine_p.change$changed %>% 
               table %>% 
               .["changed"]/2,
             " genes)"),
      paste0("unchanged\n(",
             tuci.urine_p.change$changed %>% 
               table %>% 
               .["unchanged"]/2,
             " genes)")),
    values = c("lightpink1","khaki")) +  
  xlab(NULL) +
  ylab(NULL) +
  guides(fill = "none") +
  ggtitle("p-values of differential expression\nin urine samples") +
  scale_x_discrete(
    labels = c("original\ndata",
               "processed\ndata")) +
  theme(
    axis.line = element_line(),
    panel.background = element_blank(),
    plot.title = element_text(
      size = 20,face = "bold",hjust = 0),
    axis.text.x = element_text(
      size = 20,hjust = 0.5),
    axis.text.y = element_text(
      size = 20),
    legend.title = element_blank(),
    legend.key = element_rect(
      fill = "white",color = "white"),
    legend.key.height = unit(3,"cm"),
    legend.key.width = unit(0.5,"cm"),
    legend.text = element_text(size = 20))
```

# Differential expression between matched samples

## Combine replicates

Combine replicates by averaging each gene so that each MTU patient has only one tissue and one urine sample.

```{r}
# Generate a new data frame for replicate-free expression matrix

for (.n.sample in 1:ncol(exp.proc.mtu))
{
  # check if the current sample is a replicate
  .sample.id <- 
    colnames(exp.proc.mtu)[.n.sample]
  .patient.id <-
    sample.info$PatientID[
      match(.sample.id,
            sample.info$SampleID.readable)]
  .sample.type <-
    sample.info$SampleType[
      match(.sample.id,
            sample.info$SampleID.readable)]
  .rep.all.info <-
    sample.info %>%
    dplyr::filter(SampleType == .sample.type &
                    PatientID == .patient.id)
  
  # Check and average the replicates.
  # Integrate the sample into a new dataset.
  if (.n.sample == 1 & nrow(.rep.all.info) == 1)
  {
    exp.mtu.rep_comb <- 
      data.frame(
        exp.proc.mtu[.rep.all.info$SampleID.readable])
  }
  else if (.n.sample == 1 & nrow(.rep.all.info) > 1)
  {
    exp.mtu.rep_comb <- 
      data.frame(
        exp.proc.mtu[.rep.all.info$SampleID.readable] %>%
          apply(X = .,MARGIN = 1,FUN = mean))
  }
  else if (.n.sample > 1 & nrow(.rep.all.info) == 1)
    {
    exp.mtu.rep_comb %<>%
      cbind(exp.proc.mtu[.rep.all.info$SampleID.readable])
    }
  else if (.n.sample > 1 & nrow(.rep.all.info) > 1)
    {
    if (.patient.id != .last.sample["patient.id"] |
        .sample.type != .last.sample["sample.type"])
      {
      exp.mtu.rep_comb %<>% 
        cbind(exp.proc.mtu[.rep.all.info$SampleID.readable] %>%
                apply(X = .,MARGIN = 1,FUN = mean))
    }}
  
  # Rename the new column by "patient ID + sample type"
  colnames(exp.mtu.rep_comb)[ncol(exp.mtu.rep_comb)] <-
    .rep.all.info$ID_type[1]
  
  # Save the current sample type and patient ID. If the next
  # sample is a replicate, we will skip it.
  .last.sample <- 
    c("patient.id" = .patient.id,
      "sample.type" = .sample.type)
}

rm(.n.sample,.sample.id,.patient.id,.sample.type,
   .rep.all.info,.last.sample)
```

Change the expression data into a long version.

```{r}
# replicate-combined data
exp.mtu.rep_comb.long <- 
  exp.mtu.rep_comb %>% 
  # Put gene symbols into the data frame
  dplyr::mutate(
    gene.symbol = rownames(.)) %>% 
  reshape2::melt(
    id.vars = "gene.symbol",
    variable.name = "sample.id",
    value.name = "expression") %>% 
  # Add sample type
  dplyr::mutate(
    sample.type = 
      case_when(
        str_detect(.$sample.id,"T") ~ "tissue",
        str_detect(.$sample.id,"U") ~ "urine")) %>% 
  # Add patient id
  dplyr::mutate(
    patient.id = str_extract(.$sample.id,"Q-\\d+"))

# individual sample data
exp.proc.mtu.long <- 
  exp.proc.mtu %>% 
  # Put gene symbols into the data frame
  dplyr::mutate(
    gene.symbol = rownames(.)) %>% 
  reshape2::melt(
    id.vars = "gene.symbol",
    variable.name = "sample.id",
    value.name = "expression") %>% 
  # Add patient IDs and sample types
  dplyr::left_join(
    x = .,y = mtu.sample,
    by = c("sample.id" = "SampleID.readable"))

colnames(exp.proc.mtu.long) <- 
  c("gene.symbol","sample.id","expression",
    "patient.id","sample.type")
```

## Statistical tests

Normality test

```{r}
exp.mtu.rep_comb %>% 
  t %>% 
  data.frame(check.names = F) %>% 
  shapiro_test(vars = colnames(.)) %>% 
  dplyr::arrange(p)
```

```{r}
exp.proc.mtu %>% 
  t %>% 
  data.frame(check.names = F) %>% 
  shapiro_test(vars = colnames(.)) %>% 
  dplyr::arrange(p)
```

Most genes are not normally distributed. Perform non-parametric tests.

```{r}
DE.mtu.wilcox <- 
  exp.mtu.rep_comb.long %>%
  dplyr::group_by(gene.symbol) %>%
  rstatix::wilcox_test(expression ~ sample.type,
                       paired = T) %>% 
  adjust_pvalue(p.col = "p",method = "BH",
                output.col = "p.adj")

DE.mtu.wilcox
```

There are only two genes differentially expressed between tissue and urine samples: *RPEL1* and *BHLHE22*.

## Significant genes

Extract only significant genes for visualization. Since the attributes of the extracted wilcoxon result are inherited from the original test result, it still stores all genes which influences the effect of the `add_xy_position()` function. Therefore, we extract the expression data of the significant genes and perform the test again.

```{r}
# Extract significant gene symbols
mtu.deg <- 
  DE.mtu.wilcox %>% 
  dplyr::filter(p.adj < 0.05) %>% 
  .$gene.symbol

# Extract expression data of the significant genes
exp.mtu.deg.long <- 
  exp.mtu.rep_comb.long %>% 
  dplyr::filter(gene.symbol %in% mtu.deg)

# Re-do the test
mtu.deg.wilcox <- 
  exp.mtu.deg.long %>% 
  dplyr::group_by(gene.symbol) %>%
  rstatix::wilcox_test(expression ~ sample.type,paired = T) %>% 
  add_significance(p.col = "p",output.col = "p.sig") %>% 
  add_xy_position(x = "gene.symbol",
                  dodge = 0.8,
                  step.increase = 0.1)
```

## Visualization

```{r}
ggboxplot(data = exp.mtu.deg.long,
          x = "gene.symbol",y = "expression",
          fill = "sample.type",
          xlab = F,ylab = F,size = 0.2,width = 0.4,
          bxp.errorbar = T,bxp.errorbar.width = 0.4,
          title = "differentially expressed genes\nbetween tissue and urine samples",
          ggtheme = theme(axis.line = element_line(),
                          panel.background = element_blank(),
                          plot.title =
                            element_text(size = 20,face = "bold",
                                         hjust = 0),
                          axis.text.y = element_text(size = 20),
                          axis.text.x = element_text(size = 20,
                                                      face = "italic"),
                          legend.title = element_blank(),
                          legend.key = element_rect(fill = "white"),
                          legend.key.height = unit(2,"cm"),
                          legend.key.width = unit(0.5,"cm"),
                          legend.text = element_text(size = 20))) +
  scale_fill_manual(
    labels = c(
      paste0("tissue\n(n=",mtu.deg.wilcox$n1[1],")"),
      paste0("urine\n(n=",mtu.deg.wilcox$n2[1],")")),
    values = c("orange","lightskyblue")) +
  stat_pvalue_manual(mtu.deg.wilcox,label = "p.sig",
                     tip.length = 0.01,hide.ns = T,label.size = 7)
```

These two genes are both higly expressed in urine compared to prostate tissue samples. Human Protein Atlas shows that in all organs, *BHLHE22* is only enriched in Cerebral cortex, Cerebellum and Hippocampus tissues, while *RPEL1* is enriched many organs including Kidney, urinary bladder, epididymis, and seminal vesicles. This could explain the enrichment of *RPEL1* in urine compared to prostate tissue samples.

# GSEA: Epithelial and stromal gene sets

Perform enrichment analysis of the epithelial and stromal gene sets, on tissue and urine samples

## Overall tissue and urine

Gene set data

It should be of data.frame or tibble class including two columns

1.  gene set name (e.g., epithelial gene set);

2.  gene symbol or ID included in this set. Each row represents one gene.

```{r}
GSEA.ES.geneset <- 
  rbind(
    data.frame(gene.set = "epithelial gene set",
               gene.symbol = epi.gene.set$Gene),
    data.frame(gene.set = "stromal gene set",
               gene.symbol = str.gene.set$Gene))
```

Check if the gene symbols are in the expression data

```{r}
GSEA.ES.geneset$gene.symbol %in%
  rownames(exp.proc.sample_col)
```

All gene symbols in the expression set are the same as the epithelial and stromal gene sets.

Ordered gene list

It should be of numeric class. Values should be named by the gene symbols or IDs as that in the gene set data. These values could be expression values in a single sample, or log fold changes between two groups of samples. Values should be ordered. In this part, the mean value of each gene in all tissue and all urine samples are calculated, respectively. Subsequently, the difference between the tissue mean and the urine mean is calculated in and GSEA is performed.

```{r}
exp.proc.sample_col.T_U_mean <- 
  exp.proc.sample_col

# Tissue mean
exp.proc.sample_col.T_U_mean$tissue.mean <- 
  exp.proc.sample_col.T_U_mean %>%
  dplyr::select(contains("T")) %>% 
  apply(X = .,MARGIN = 1,FUN = mean)

# Urine mean
exp.proc.sample_col.T_U_mean$urine.mean <- 
  exp.proc.sample_col.T_U_mean %>%
  dplyr::select(contains("U")) %>% 
  apply(X = .,MARGIN = 1,FUN = mean)

# Mean difference
exp.proc.sample_col.T_U_mean %<>% 
  dplyr::mutate(
    mean.diff = tissue.mean-urine.mean)

# Order genes by the mean difference
exp.proc.sample_col.T_U_mean %<>%
  dplyr::arrange(desc(mean.diff))

# Extract the ordered gene list and the mean difference value
GSEA.ord_val.mean_diff.all_sam <- 
  exp.proc.sample_col.T_U_mean$mean.diff

names(GSEA.ord_val.mean_diff.all_sam) <- 
  rownames(exp.proc.sample_col.T_U_mean)
```

Run GSEA

```{r}
GSEA.ES.all_sam.result <- 
  GSEA(geneList = GSEA.ord_val.mean_diff.all_sam,
       TERM2GENE = GSEA.ES.geneset,
       pAdjustMethod = "none",by = "fgsea",
       pvalueCutoff = 1,verbose = T) %>%
  suppressWarnings
```

Visualize

```{r cache=TRUE}
gseaplot2(GSEA.ES.all_sam.result,
          geneSetID = 1:2,pvalue_table = TRUE,
          title = "epithelial & stromal gene set enrichment in prostate tissue vs urine",
          base_size = 18)
```

In the whole dataset, the epithelial gene set is more enriched in urine than tissue, while the stromal gene set does not show significant enrichment in tissue and urine.

# GSVA

Gene set Variation Analysis (GSVA) is performed to calculate enrichment scores. GSVA can transform a "sample by gene" matrix to a "sample by gene set" matrix, where each value is the enrichment scores of a specific gene set in a specific sample.

## Organize gene sets

GSVA requires a list of gene sets (list class). Each gene set is a character vector of gene symbols or ids (depends on the expression matrix). The name of each vector is the gene set name.

Epithelial and stromal gene sets.

```{r}
GSVA.geneset.ES <- 
  list("epithelial gene set" = epi.gene.set$Gene,
       "stromal gene set" = str.gene.set$Gene)
```

MSigDB gene sets

Retrieve MSigDB gene sets.

```{r}
msigdbr.geneset.raw <-
  msigdbr(species = "Homo sapiens")
```

Re-organize the gene sets.

```{r}
# For GSVA: an input list
GSVA.geneset.msigdbr <- 
  msigdbr.geneset.raw %>% 
  # Split the genes into a gene set list
  split(x = .$gene_symbol,
        f = .$gs_id) %>% 
  # Remove duplicate genes
  lapply(unique)

# For gene set filtration: a non-redundant gene set
# data frame
msigdbr.geneset.unique <- 
  msigdbr.geneset.raw %>% 
  dplyr::select(
    c("gs_id","gs_name","gs_collection",
      "gs_collection_name")) %>%
  unique %>% 
  dplyr::arrange(gs_collection,gs_id)
```

Check if all gene symbols are in the MSigDB gene sets.

```{r}
gene.unmatched <- 
  data.frame(
    gene.alias =
      rownames(exp.proc.sample_col) %>% 
      .[! . %in% msigdbr.geneset.raw$gene_symbol]) %>% 
  dplyr::mutate(
    gene.id = 
      gene.id$ID[
        match(.$gene.alias,gene.id$Gene)]) %>% 
  dplyr::arrange(gene.id)

gene.unmatched
```

There are 12 gene symbols not in the gene set. Use Entrez IDs to re-check them.

```{r}
# Retrieve the database
ensembl <- 
  useEnsembl(biomart = "genes",
             dataset = "hsapiens_gene_ensembl")

# Find and add the gene symbols 
gene.unmatched <-
  getBM(attributes = c("entrezgene_id", "hgnc_symbol"),
        filters = "entrezgene_id",
        values = gene.unmatched$gene.id,
        mart = ensembl) %>% 
  merge(x = gene.unmatched,
        y = .,
        by.x = "gene.id",
        by.y = "entrezgene_id")

# Change these gene symbols in the expression matrix
rownames(exp.proc.sample_col) %<>%
  replace(x = .,
          list = . %in% gene.unmatched$gene.alias,
          values = gene.unmatched$hgnc_symbol[
            match(.,gene.unmatched$gene.alias) %>%
              na.omit()])

colnames(exp.ori.sample_row) %<>%
  replace(x = .,
          list = . %in% gene.unmatched$gene.alias,
          values = gene.unmatched$hgnc_symbol[
            match(.,gene.unmatched$gene.alias) %>%
              na.omit()])

# Reorder rows
exp.proc.sample_col %<>% 
  .[order(rownames(.)),]

exp.ori.sample_row %<>% 
  .[,order(colnames(.))]

# Add the symbols to the gene id pool
gene.id <- 
  rbind(
    gene.id,
    data.frame(
      Gene = gene.unmatched$hgnc_symbol,
      ID = gene.unmatched$gene.id)) %>% 
  unique()

table(rownames(exp.proc.sample_col) %in%
        msigdbr.geneset.raw$gene_symbol)
```

All gene symbols are in the MSigDB.

Check if gene symbols of the epithelial and stromal gene sets need to be changed.

```{r}
epi.gene.set$ID %in% gene.unmatched$gene.id
```

```{r}
str.gene.set$ID %in% gene.unmatched$gene.id
```

Gene symbols of the epithelial and stromal gene sets are not in the unmatched genes and thus do not need to be changed.

## Organize expression matrix

GSVA requires an expression matrix (matrix class). Rows are genes and rownames should be gene symbols or ids. Columns are samples and column names are sample ids.

Processed data

```{r}
GSVA.expset <-
  exp.proc.sample_col %>% 
  as.matrix()
```

Original data

```{r}
# tissue samples
GSVA.expset.ori.tissue <- 
  exp.ori.sample_row %>% 
  .[str_detect(rownames(.),"T"),] %>% 
  t

# urine samples
GSVA.expset.ori.urine <- 
  exp.ori.sample_row %>% 
  .[str_detect(rownames(.),"U"),] %>% 
  t
```

## Perform GSVA

For a gene set in a sample, GSVA mainly includes several steps:

1.  Arrange all genes in ascending order.
2.  Start two independent random walks from the gene with the lowest expression.
    -   Positive random walk. When the algorithm meets a gene in the gene set, update the Empirical Cumulative Distribution Function (ECDF) of the in-set genes. It is equal to total number of met in-set genes / size of the gene set.

    -   Negative random walk. When the algorithm meets a gene not in the gene set, update the ECDF of the out-of-set genes. It is equal to total number of met out-of-set genes / size of the whole gene list.
3.  Depict the trend of the two ECDFs. Calculate the Enrichment Score (ES) based on the deviation of the ECDFs (see below).

The function `gsva()` requires a parameter set organized by the function `gsvaParam()` . The latter function has some important parameters:

-   `kcdf`: Character vector of length 1 denoting the kernel to use during the non-parametric estimation of the empirical cumulative distribution function (ECDF) of expression levels across samples. Choose "Poisson" for integer value from raw RNA-Seq data and "Gaussian" for log-transformed data. "auto" forces the function to detects the data automatically. "none" prevents the function from using the kernel approach to estimate the ECDF.

-   `maxDiff` : logical value. If `FALSE`, ES is calculated as the maximum distance of the random walk from 0. The ES shows the extent of the expression change rather than the direction of the change (*i.e.*, enrichment). If `TRUE` , ES is calculated as the magnitude difference between the largest positive and negative random walk deviations. The symbol of the ES represents the the direction of the change. Default is TRUE.

-   `absRanking` : logical value. Only works when `maxDiff=TRUE` . If `FALSE` , ES = \|largest positive random walk deviation\| - \|largest negative random walk deviation\|. The symbol of the ES represents the direction of the change. If `TRUE` , ES = \|largest positive random walk deviation\| + \|largest negative random walk deviation\|. In this scenario, ES is always positive. Default is FALSE.

-   `minSize` : the threshold of filtering gene sets by the number of overlapping genes between the expression matrix and the gene set.

### Processed data: Epithelial and stromal gene sets

```{r}
GSVA.result.ES.all_sam <- 
  gsvaParam(
    exprData = GSVA.expset,
    geneSets = GSVA.geneset.ES,
    kcdf = "none",
    maxDiff = T,
    absRanking = T,
    minSize = 1) %>%
  gsva(verbose = T) %>% 
  data.frame(check.names = F)
```

### Processed data: MSigDB gene sets

To narrow down the gene set list, we focus only on:

-   C4: C4 is the collection of computational gene sets defined by mining large collections of cancer-oriented expression data.

-   C5 Cellular Component: Gene sets that contain genes annotated by the same cellular component ontology term.

-   C6: Gene sets that represent signatures of cellular pathways which are often dis-regulated in cancer.

-   C7 ImmuneSigDB: Gene sets representing chemical and genetic perturbations of the immune system generated by manual curation of published studies in human and mouse immunology.

```{r}
overlap.gene.num <- 10

GSVA.result.msigdb.all_sam <- 
  gsvaParam(
    exprData = GSVA.expset,
    geneSets = GSVA.geneset.msigdbr,
    kcdf = "none",
    maxDiff = T,
    absRanking = T,
    minSize = overlap.gene.num) %>%
  gsva(verbose = T) %>% 
  data.frame(check.names = F)

# Extract specific gene sets
GSVA.result.msigdb.all_sam <- 
  GSVA.result.msigdb.all_sam %>% 
  ## Add category information of the gene sets
  dplyr::mutate(
    gene.set = rownames(.),
    cat = 
      msigdbr.geneset.unique$gs_collection[
        match(rownames(.),
              msigdbr.geneset.unique$gs_id)],
    subcat = 
      msigdbr.geneset.unique$gs_collection_name[
        match(rownames(.),
              msigdbr.geneset.unique$gs_id)]) %>% 
  ## Filter gene sets
  dplyr::filter(
    (cat == "C7" & subcat == "ImmuneSigDB"))
```

### Original data: epithelial and stromal gene sets

```{r}
# tissue samples
GSVA.result.ES.ori_tissue <- 
  gsvaParam(
    exprData = GSVA.expset.ori.tissue,
    geneSets = GSVA.geneset.ES,
    kcdf = "none",
    maxDiff = T,
    absRanking = T,
    minSize = 1) %>%
  gsva(verbose = T) %>% 
  data.frame(check.names = F)

# urine samples
GSVA.result.ES.ori_urine <- 
  gsvaParam(
    exprData = GSVA.expset.ori.urine,
    geneSets = GSVA.geneset.ES,
    kcdf = "none",
    maxDiff = T,
    absRanking = T,
    minSize = 1) %>%
  gsva(verbose = T) %>% 
  data.frame(check.names = F)
```

## Annotations for visualization

Sample annotations

```{r}
GSVA.anno.sam <- 
  data.frame(row.names = 
               colnames(GSVA.result.msigdb.all_sam),
             "sample type"= 
               rep("",ncol(GSVA.result.msigdb.all_sam)),
             check.names = F) %>% 
  ## Add sample type and patient ID information
  dplyr::mutate(
    `sample type` = sample.info$SampleType[
      match(rownames(.),sample.info$SampleID.readable)],
    `patient ID` = sample.info$PatientID[
      match(rownames(.),sample.info$SampleID.readable)]) %>% 
  ## Order samples by sample type and then patient ID
  dplyr::arrange(`sample type`,`patient ID`)
```

Gene set annotations

```{r}
GSVA.anno.geneset.msigdb <- 
  msigdbr.geneset.unique %>% 
  dplyr::filter(gs_id %in%
                  rownames(GSVA.result.msigdb.all_sam))
```

## Epithelial and stromal genes

Extract gene set results

```{r}
# MTU samples
GSVA.result.ES.mtu <- 
  GSVA.result.ES.all_sam %>% 
  .[,colnames(.) %in% mtu.sample$SampleID.readable]

# All samples: original data
GSVA.result.ES.all_sam_ori <- 
  cbind(
    GSVA.result.ES.ori_tissue,
    GSVA.result.ES.ori_urine)
```

Combine replicates by averaging each gene set so that each MTU patient has only one tissue and one urine sample.

```{r}
# Generate a new data frame for replicate-free enrichment matrix
for (.n.sample in 1:ncol(GSVA.result.ES.mtu))
{
  # check if the current sample is a replicate
  .sample.id <- 
    colnames(GSVA.result.ES.mtu)[.n.sample]
  .patient.id <-
    sample.info$PatientID[
      match(.sample.id,
            sample.info$SampleID.readable)]
  .sample.type <-
    sample.info$SampleType[
      match(.sample.id,
            sample.info$SampleID.readable)]
  .rep.all.info <-
    sample.info %>%
    dplyr::filter(SampleType == .sample.type &
                    PatientID == .patient.id)
  
  # Check and average the replicates.
  # Integrate the sample into a new dataset.
  if (.n.sample == 1 & nrow(.rep.all.info) == 1)
  {
    GSVA.result.ES.rep_comb <- 
      data.frame(
        GSVA.result.ES.mtu[
          .rep.all.info$SampleID.readable])
  }
  else if (.n.sample == 1 & nrow(.rep.all.info) > 1)
  {
    GSVA.result.ES.rep_comb <- 
      data.frame(
        GSVA.result.ES.mtu[
          .rep.all.info$SampleID.readable] %>%
          apply(X = .,MARGIN = 1,FUN = mean))
  }
  else if (.n.sample > 1 & nrow(.rep.all.info) == 1)
    {
    GSVA.result.ES.rep_comb %<>%
      cbind(
        GSVA.result.ES.mtu[
          .rep.all.info$SampleID.readable])
    }
  else if (.n.sample > 1 & nrow(.rep.all.info) > 1)
    {
    if (.patient.id != .last.sample["patient.id"] |
        .sample.type != .last.sample["sample.type"])
      {
      GSVA.result.ES.rep_comb %<>% 
        cbind(
          GSVA.result.ES.mtu[
            .rep.all.info$SampleID.readable] %>%
            apply(X = .,MARGIN = 1,FUN = mean))
    }}
  
  # Rename the new column by "patient ID + sample type"
  colnames(GSVA.result.ES.rep_comb)[
    ncol(GSVA.result.ES.rep_comb)] <-
    .rep.all.info$ID_type[1]
  
  # Save the current sample type and patient ID. If the next
  # sample is a replicate, we will skip it.
  .last.sample <- 
    c("patient.id" = .patient.id,
      "sample.type" = .sample.type)
}

rm(.n.sample,.sample.id,.patient.id,.sample.type,
   .rep.all.info,.last.sample)
```

Change enrichment score matrices into a long version

```{r}
# All samples: processed data
GSVA.result.ES.all_sam.long <- 
  GSVA.result.ES.all_sam %>% 
  t %>% 
  data.frame(check.names = F) %>% 
  ## Add sample ID and type annotations
  dplyr::mutate(
    sample.id = rownames(.),
    sample.type =
      GSVA.anno.sam$`sample type`[
        match(rownames(.),
              rownames(GSVA.anno.sam))]) %>% 
  reshape2::melt(
    id.vars = c("sample.id","sample.type"),
    variable.name = "gene.set",
    value.name = "enrichment.score")

# Replicate-combined data. We perform comparison
# in: epithelial vs stromal gene sets, and
# tissue vs urine.
GSVA.result.ES.rep_comb.long <- 
  GSVA.result.ES.rep_comb %>% 
  t %>% 
  data.frame(check.names = F) %>% 
  ## Add sample ID, type annotations, and patient ID
  dplyr::mutate(
    sample.id = rownames(.)) %>% 
  dplyr::mutate(
    sample.type = 
      case_when(
        str_detect(.$sample.id,"T") ~ "tissue",
        str_detect(.$sample.id,"U") ~ "urine"),
    patient.id =
      str_extract(.$sample.id,"Q-\\d+")) %>% 
  # Change to a long version
  reshape2::melt(
    id.vars = c("sample.id","sample.type","patient.id"),
    variable.name = "gene.set",
    value.name = "enrichment.score") %>% 
  # Compare the enrichment scores in tissue vs urine and 
  # add a column to record the result. For the convenience 
  # of processing, jaxtapose the enrichment scores in
  # tissue and urine first.
  reshape2::dcast(
    patient.id + gene.set ~ sample.type,
    value.var = "enrichment.score") %>% 
  # Add a result column
  dplyr::mutate(
    tissue.vs.urine = 
      if_else(tissue > urine,
              "higher in tissue",
              "higher in urine")) %>% 
  # Change the data again to the long version
  reshape2::melt(
    id.vars = c("patient.id","gene.set","tissue.vs.urine"),
    measure.vars = c("tissue","urine"),
    variable.name = "sample.type",
    value.name = "enrichment.score") %>% 
  # Add x ordinates for ggplot
  dplyr::mutate(
    geom.line.x = 
      case_when(
        gene.set == "epithelial gene set" &
          sample.type == "tissue" ~ 0.8,
        gene.set == "epithelial gene set" &
          sample.type == "urine" ~ 1.2,
        gene.set == "stromal gene set" &
          sample.type == "tissue" ~ 1.8,
        gene.set == "stromal gene set" &
          sample.type == "urine" ~ 2.2))

# All samples: original data
GSVA.result.ES.all_sam_ori.long <- 
  GSVA.result.ES.all_sam_ori %>% 
  t %>% 
  data.frame(check.names = F) %>% 
  ## Add sample ID and type annotations
  dplyr::mutate(
    sample.id = rownames(.),
    sample.type =
      GSVA.anno.sam$`sample type`[
        match(rownames(.),
              rownames(GSVA.anno.sam))]) %>% 
  reshape2::melt(
    id.vars = c("sample.id","sample.type"),
    variable.name = "gene.set",
    value.name = "enrichment.score")
```

### Normality test

All samples: processed

```{r}
GSVA.result.ES.all_sam.long %>% 
  group_by(sample.type,gene.set) %>% 
  shapiro_test(vars = "enrichment.score")
```

MTU samples

```{r}
GSVA.result.ES.rep_comb.long %>% 
  group_by(sample.type,gene.set) %>% 
  shapiro_test(vars = "enrichment.score")
```

All samples: original

```{r}
GSVA.result.ES.all_sam_ori.long %>% 
  group_by(sample.type,gene.set) %>% 
  shapiro_test(vars = "enrichment.score")
```

Enrichment scores of the two gene sets in tissue and urine samples are all normally distributed.

### t-test

MTU samples: tissue vs urine

```{r}
GSVA.stat.ES.rep_comb.t_vs_u <-
  GSVA.result.ES.rep_comb.long %>%
  group_by(gene.set) %>% 
  rstatix::t_test(
    enrichment.score ~ sample.type,
    paired = T) %>% 
  add_significance() %>% 
  add_xy_position(
    x = "gene.set",dodge = 0.8) %>% 
  # For visualization, increase y.position more
  dplyr::mutate(
    y.position = y.position + 0.05
  )

GSVA.stat.ES.rep_comb.t_vs_u
```

MTU samples: epithelial vs stromal gene sets

```{r}
GSVA.result.ES.rep_comb.long %>%
  group_by(sample.type) %>% 
  rstatix::t_test(
    enrichment.score ~ gene.set,
    paired = T)
```

All processed samples: tissue vs urine

```{r}
GSVA.stat.ES.all_sam.t_vs_u <- 
  GSVA.result.ES.all_sam.long %>%
  group_by(gene.set) %>% 
  rstatix::t_test(
    enrichment.score ~ sample.type) %>% 
  add_significance() %>% 
  add_xy_position(
    x = "gene.set",dodge = 0.8) %>% 
  # For visualization, increase y.position more
  dplyr::mutate(
    y.position = y.position + 0.05)

GSVA.stat.ES.all_sam.t_vs_u
```

All processed samples: epithelial vs stromal gene set

```{r}
GSVA.stat.ES.all_sam.e_vs_s <- 
  GSVA.result.ES.all_sam.long %>%
  group_by(sample.type) %>% 
  rstatix::t_test(
    enrichment.score ~ gene.set,
    paired = T) %>% 
  add_xy_position(x = "sample.type",dodge = 0.8) %>% 
  # For visualization, increase y.position more
  dplyr::mutate(
    y.position = y.position + 0.05)

GSVA.stat.ES.all_sam.e_vs_s
```

All original samples: epithelial vs stromal gene set

```{r}
GSVA.stat.ES.all_sam_ori.e_vs_s <- 
  GSVA.result.ES.all_sam_ori.long %>%
  group_by(sample.type) %>% 
  rstatix::t_test(
    enrichment.score ~ gene.set,
    paired = T) %>% 
  add_xy_position(x = "sample.type",dodge = 0.8) %>% 
  # For visualization, increase y.position more
  dplyr::mutate(
    y.position = y.position + 0.05)

GSVA.stat.ES.all_sam_ori.e_vs_s
```

### Visualization

Tissue vs urine: MTU samples

```{r fig.height=4, fig.width=6, cache=TRUE}
ggplot(
  data = GSVA.result.ES.rep_comb.long,
  mapping = aes(
    x = gene.set,y = enrichment.score,
    fill = sample.type)) +
  stat_summary(
    fun.min = function(x){quantile(x)["75%"]+1.5*sd(x)},
    fun.max = function(x){quantile(x)["25%"]-1.5*sd(x)},
    geom = "errorbar",
    position = position_dodge(width = 0.8),
    width = 0.4,color = "black",linewidth = 0.5,) +
  geom_boxplot(
    position = position_dodge(0.8)) +
  scale_fill_manual(
    labels = c(
      paste0("tissue\n(n=",
             GSVA.stat.ES.rep_comb.t_vs_u$n1[1],")"),
      paste0("urine\n(n=",
             GSVA.stat.ES.rep_comb.t_vs_u$n2[1],")")),
    values = c("darkmagenta","yellow3")) +
  geom_point(
    position = position_dodge(0.8)) +
  geom_line(
    data = 
      GSVA.result.ES.rep_comb.long %>% 
      dplyr::filter(gene.set == "epithelial gene set"),
    mapping = aes(
      x = geom.line.x,y = enrichment.score,
      group = patient.id,color = tissue.vs.urine)) +
  geom_line(
    data = 
      GSVA.result.ES.rep_comb.long %>% 
      dplyr::filter(gene.set == "stromal gene set"),
    mapping = aes(
      x = geom.line.x,y = enrichment.score,
      group = patient.id,color = tissue.vs.urine)) +
  scale_color_manual(
    labels = c("higher in tissue",
               "higher in urine"),
    values = c("dodgerblue","sienna")) +
  xlab(NULL) +
  ylab(NULL) +
  ggtitle("enrichment scores\nin MTU samples") +
  scale_x_discrete(
    labels = c("epithelial\ngene set",
               "stromal\ngene set")) +
  stat_pvalue_manual(
    GSVA.stat.ES.rep_comb.t_vs_u,
    label = "p.signif",
    tip.length = 0.01,label.size = 7) +
  theme(
    axis.line = element_line(),
    panel.background = element_blank(),
    plot.title = element_text(
      size = 20,face = "bold",hjust = 0),
    axis.text.x = element_text(
      size = 20,hjust = 0.5),
    axis.text.y = element_text(
      size = 20),
    legend.title = element_blank(),,
    legend.key = element_rect(
      fill = "white",color = "white"),
    legend.key.height = unit(1,"cm"),
    legend.key.width = unit(0.5,"cm"),
    legend.text = element_text(size = 20))
```

Tissue vs urine: all samples

```{r fig.height=4, fig.width=6, cache=TRUE}
ggplot(
  data = GSVA.result.ES.all_sam.long,
  mapping = aes(
    x = gene.set,y = enrichment.score,
    fill = sample.type)) +
  stat_summary(
    fun.min = function(x){quantile(x)["75%"]+1.5*sd(x)},
    fun.max = function(x){quantile(x)["25%"]-1.5*sd(x)},
    geom = "errorbar",
    position = position_dodge(width = 0.8),
    width = 0.4,color = "black",linewidth = 0.5) +
  geom_boxplot(
    position = position_dodge(0.8),
    coef = 0,
    outlier.shape = NA) +
  scale_fill_manual(
    labels = c(
      paste0("tissue\n(n=",
             GSVA.stat.ES.all_sam.t_vs_u$n1[1],")"),
      paste0("urine\n(n=",
             GSVA.stat.ES.all_sam.t_vs_u$n2[1],")")),
    values = c("darkmagenta","yellow3")) +
  xlab(NULL) +
  ylab(NULL) +
  ggtitle("enrichment scores\nin all samples\ntissue vs urine") +
  scale_x_discrete(
    labels = c("epithelial\ngene set",
               "stromal\ngene set")) +
  stat_pvalue_manual(
    GSVA.stat.ES.all_sam.t_vs_u,
    label = "p.signif",
    tip.length = 0.01,label.size = 7) +
  theme(
    axis.line = element_line(),
    panel.background = element_blank(),
    plot.title = element_text(
      size = 20,face = "bold",hjust = 0),
    axis.text.x = element_text(
      size = 20,hjust = 0.5),
    axis.text.y = element_text(
      size = 20),
    legend.title = element_blank(),,
    legend.key = element_rect(
      fill = "white",color = "white"),
    legend.key.height = unit(1,"cm"),
    legend.key.width = unit(0.5,"cm"),
    legend.text = element_text(size = 20))
```

Epithelial vs stromal: processed data of all samples

```{r fig.height=4, fig.width=6, cache=TRUE}
ggplot(
  data = GSVA.result.ES.all_sam.long,
  mapping = aes(
    x = sample.type,y = enrichment.score,
    fill = gene.set)) +
  stat_summary(
    fun.min = function(x){quantile(x)["75%"]+1.5*sd(x)},
    fun.max = function(x){quantile(x)["25%"]-1.5*sd(x)},
    geom = "errorbar",
    position = position_dodge(width = 0.8),
    width = 0.4,color = "black",linewidth = 0.5) +
  geom_boxplot(
    position = position_dodge(0.8),
    coef = 0,
    outlier.shape = NA) +
  scale_fill_manual(
    labels = c("epithelial\ngene set",
               "stromal\ngene set"),
    values = c("lightcoral","mediumturquoise")) +
  xlab(NULL) +
  ylab(NULL) +
  ggtitle("enrichment scores\nin all samples of processed data\nepithelial vs stromal gene set") +
  scale_x_discrete(
    labels = c(
      paste0("tissue\n(n=",
             GSVA.stat.ES.all_sam.e_vs_s$n1[1],")"),
      paste0("urine\n(n=",
             GSVA.stat.ES.all_sam.e_vs_s$n1[2],")"))) +
  xlab(NULL) +
  ylab(NULL) +
  stat_pvalue_manual(
    GSVA.stat.ES.all_sam.e_vs_s,
    label = "p",
    tip.length = 0.01,label.size = 7) +
  theme(
    axis.line = element_line(),
    panel.background = element_blank(),
    plot.title = element_text(
      size = 20,face = "bold",hjust = 0),
    axis.text.x = element_text(
      size = 20,hjust = 0.5),
    axis.text.y = element_text(
      size = 20),
    legend.title = element_blank(),,
    legend.key = element_rect(
      fill = "white",color = "white"),
    legend.key.height = unit(1,"cm"),
    legend.key.width = unit(0.5,"cm"),
    legend.text = element_text(size = 20))
```

Epithelial vs stromal: original data of all samples

```{r fig.height=4, fig.width=6, cache=TRUE}
ggplot(
  data = GSVA.result.ES.all_sam_ori.long,
  mapping = aes(
    x = sample.type,y = enrichment.score,
    fill = gene.set)) +
  stat_summary(
    fun.min = function(x){quantile(x)["75%"]+1.5*sd(x)},
    fun.max = function(x){quantile(x)["25%"]-1.5*sd(x)},
    geom = "errorbar",
    position = position_dodge(width = 0.8),
    width = 0.4,color = "black",linewidth = 0.5) +
  geom_boxplot(
    position = position_dodge(0.8),
    coef = 0,
    outlier.shape = NA) +
  scale_fill_manual(
    labels = c("epithelial\ngene set",
               "stromal\ngene set"),
    values = c("lightcoral","mediumturquoise")) +
  xlab(NULL) +
  ylab(NULL) +
  ggtitle("enrichment scores\nin all samples of original data\nepithelial vs stromal gene set") +
  scale_x_discrete(
    labels = c(
      paste0("tissue\n(n=",
             GSVA.stat.ES.all_sam_ori.e_vs_s$n1[1],")"),
      paste0("urine\n(n=",
             GSVA.stat.ES.all_sam_ori.e_vs_s$n1[2],")"))) +
  xlab(NULL) +
  ylab(NULL) +
  stat_pvalue_manual(
    GSVA.stat.ES.all_sam_ori.e_vs_s,
    label = "p",
    tip.length = 0.01,label.size = 7) +
  theme(
    axis.line = element_line(),
    panel.background = element_blank(),
    plot.title = element_text(
      size = 20,face = "bold",hjust = 0),
    axis.text.x = element_text(
      size = 20,hjust = 0.5),
    axis.text.y = element_text(
      size = 20),
    legend.title = element_blank(),,
    legend.key = element_rect(
      fill = "white",color = "white"),
    legend.key.height = unit(1,"cm"),
    legend.key.width = unit(0.5,"cm"),
    legend.text = element_text(size = 20))
```

### Summary

When comparing between tissue and urine samples, the epithelial gene set is more enriched in urine compared to tissue samples. In contrast, the stromal gene set is comparably enriched in tissue and urine samples, showing no significant differences.

When comparing between the epithelial and stromal gene sets, there are no significant differences in either tissue or urine samples. In tissue samples, the stromal gene set is more enriched than the epithelial gene set, while the p-value is marginal (0.07) and does not reach the significance. Notably, enrichment of the epithelial and stromal gene sets is also compared in the original data. The result is consistent with the comparison in the processed data, suggesting that the biological variation within the same sample type is kept.

Altogether, these results suggest that gene transcripts from the epithelial tissue exhibit a higher propensity to shed from tissue and become enriched in urine than gene transcripts from the stromal tissue.

## MSigDB gene sets

Change to the long version.

```{r}
GSVA.result.msigdb.all_sam.long <- 
  GSVA.result.msigdb.all_sam %>% 
  reshape2::melt(
    id.vars = c("gene.set","cat","subcat"),
    variable.name = "sample.id",
    value.name = "enrichment.score") %>% 
  ## Add sample type annotations
  dplyr::mutate(
    sample.type =
      GSVA.anno.sam$`sample type`[
        match(.$sample.id,
              rownames(GSVA.anno.sam))])
```

### Normality test

All samples

```{r}
GSVA.stat_norm.msigdb.all_sam <- 
  GSVA.result.msigdb.all_sam.long %>% 
  group_by(sample.type,gene.set) %>% 
  shapiro_test(vars = "enrichment.score")
```

Find gene sets not normally distributed.

```{r}
GSVA.stat_norm.msigdb.all_sam <- 
  GSVA.stat_norm.msigdb.all_sam %>% 
  dplyr::mutate(
    norm.dist = p > 0.05)

msigdb.not_norm_dis <-
  GSVA.stat_norm.msigdb.all_sam %>% 
  dplyr::filter(
    !norm.dist) %>% 
  .$gene.set %>% 
  as.character() %>% 
  unique()

length(msigdb.not_norm_dis)

GSVA.result.msigdb.all_sam.long$gene.set %>% 
  unique %>% 
  length()
```

There are 171/1055 gene sets with at least one sample group (tissue or urine) not normally distributed.

### wilcoxon test

```{r}
GSVA.stat.msigdb.all_sam.t_vs_u <- 
  GSVA.result.msigdb.all_sam.long %>%
  group_by(gene.set) %>% 
  rstatix::wilcox_test(
    enrichment.score ~ sample.type) %>% 
  adjust_pvalue(
    method = "BH",
    output.col = "p.adj") %>% 
  add_significance(
    p.col = "p.adj")
```

```{r}
GSVA.stat.msigdb.all_sam.t_u_diff <- 
  GSVA.stat.msigdb.all_sam.t_vs_u %>% 
  dplyr::filter(
    p.adj < 0.1)

nrow(GSVA.stat.msigdb.all_sam.t_u_diff)
```

There are / gene sets with significant differences between tissue and urine samples.

### Organize results

Enrichment score matrix

```{r}
# Set the number of edge gene sets (top n gene sets in 
# tissue or urine)
edge.gs.num <- 10

# Organize the matrix with annotations
GSVA.result.msigdb.t_u_diff.all_sam <- 
  ## Extract differential gene sets from the matrix
  GSVA.result.msigdb.all_sam %>% 
  dplyr::filter(
    rownames(.) %in% 
      GSVA.stat.msigdb.all_sam.t_u_diff$gene.set) %>% 
  ## Order columns by sample type and ID.
  ## Use the order of GSVA.anno.sam which was performed previously.
  dplyr::select(
    rownames(GSVA.anno.sam),"gene.set","cat","subcat") %>% 
  ## Calculate mean enrichment scores in tissue and urine 
  dplyr::mutate(
    mean.tissue = 
      .[,str_detect(string = colnames(.),
                    pattern = "T")] %>% 
      apply(1,mean),
    mean.urine =
      .[,str_detect(string = colnames(.),
                    pattern = "U")] %>% 
      apply(1,mean)) %>% 
  ## Calculate mean differences and order gene sets by
  ## this value
  dplyr::mutate(
    lfc.t_u =
      log2(mean.tissue/mean.urine)) %>% 
  dplyr::arrange(
    dplyr::desc(lfc.t_u)) %>% 
  ## Extract only edge gene sets
  dplyr::slice(
    c(1:edge.gs.num,
      (n() - edge.gs.num + 1):n())) %>%
  ## Add annotations: tissue enriched / urine enriched
  dplyr::mutate(
    direction = case_when(
      lfc.t_u > 0 ~ "tissue enriched",
      lfc.t_u < 0 ~ "urine enriched"))
  
# Extract only the enrichment score matrix
GSVA.mat.msigdb.t_u_diff.all_sam <- 
  GSVA.result.msigdb.t_u_diff.all_sam %>% 
  dplyr::mutate(
    gene.set = NULL,
    cat = NULL,
    subcat = NULL,
    mean.tissue = NULL,
    mean.urine = NULL,
    lfc.t_u = NULL,
    direction = NULL)
```

Gene set information

```{r}
# all gene sets
GSVA.anno.geneset.msigdb.t_u_diff <- 
  GSVA.anno.geneset.msigdb %>% 
  dplyr::filter(
    gs_id %in% 
      rownames(GSVA.result.msigdb.t_u_diff.all_sam)) %>% 
  ## Put gene set IDs into row names
  data.frame(
    check.names = F,
    row.names = .$gs_id) %>% 
  dplyr::mutate(
    gs_id = NULL) %>% 
  ## Curate category labels
  dplyr::mutate(
    `gene set category` = case_when(
      gs_collection %in% c("C4","C6") ~ "cancer related",
      gs_collection_name == "GO Cellular Component" ~ 
        "Gene Ontology--Cellular Component",
      gs_collection_name == "GO Molecular Function" ~ 
        "Gene Ontology--Molecular Function",
      gs_collection == "C7" ~ "immune related",
      gs_collection == "C8" ~ "cell type signature"),
    enrichment = 
      GSVA.result.msigdb.t_u_diff.all_sam$direction[
        match(
          rownames(.),
          rownames(GSVA.result.msigdb.t_u_diff.all_sam))])

# tissue enriched gene sets
GSVA.anno.geneset.msigdb.t_enrich <- 
  GSVA.anno.geneset.msigdb %>% 
  dplyr::filter(
    gs_id %in% 
      rownames(
        GSVA.result.msigdb.t_u_diff.all_sam %>% 
          dplyr::filter(direction == "tissue enriched")))

# urine enriched gene sets
GSVA.anno.geneset.msigdb.u_enrich <- 
  GSVA.anno.geneset.msigdb %>% 
  dplyr::filter(
    gs_id %in% 
      rownames(
        GSVA.result.msigdb.t_u_diff.all_sam %>% 
          dplyr::filter(direction == "urine enriched")))

GSVA.anno.geneset.msigdb.t_u_diff$gs_collection %>% 
  table
GSVA.anno.geneset.msigdb.t_enrich$gs_collection %>% 
  table
GSVA.anno.geneset.msigdb.u_enrich$gs_collection %>% 
  table
```

### Visualization

Heatmap of edge gene sets

```{r fig.height=8, fig.width=15}
pheatmap(
  GSVA.mat.msigdb.t_u_diff.all_sam,
  main = "enrichment score of MSigDB gene sets\nwith top&bottom mean differences\nin tissue vs urine samples",
  show_rownames = F,show_colnames = F,
  cluster_rows = F,cluster_cols = F,
  angle_col = 315,fontsize = 20,
  annotation_col = GSVA.anno.sam["sample type"],
  annotation_names_col = T,
  annotation_row = 
    GSVA.anno.geneset.msigdb.t_u_diff[
      c("gene set category","enrichment")],
  annotation_names_row = T,
  annotation_colors = list(
    "sample type" = c(
      "tissue" = "darkmagenta",
      "urine"="yellow3"),
    "gene set category" = c(
      "cancer related" = "indianred1",
      "Gene Ontology--Cellular Component" = "limegreen",
      "immune related" = "grey80"),
    "enrichment" = c(
      "tissue enriched" = "dodgerblue",
      "urine enriched" = "sienna"
    )))
```

### Summary

Among all MSigDB gene sets, C8 cell signature gene sets make more sense for tissue vs urine comparison. However, this collection of gene sets is limited to several projects and does not cover all cell types. In addition, C5 Gene Ontology Cellular Component gene sets classify genes based on the location of their protein or ncRNA products, rather than mRNA. To compare the enrichment of different cell types in tissue vs urine, we will use cell marker databases.

# Sample and patient info

```{r}
patient.info <- 
  sample.info %>% 
  dplyr::select("PatientID","GS","match") %>% 
  unique

patient.info$match %>% table

patient.info$GS %>% table
```

```{r}
sample.tissue <- 
  sample.info %>% 
  dplyr::filter(
    SampleType == "tissue")

sample.urine <- 
  sample.info %>% 
  dplyr::filter(
    SampleType == "urine")

sample.tissue$match %>% table
sample.urine$match %>% table
```
