---
title: "9.2. tissue vs urine re-comparison--cell markers"
author: "YD-X"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    toc: true
    number_sections: true
    latex_engine: xelatex
  html_document: 
    toc: true
    number_sections: true
---

This rmd file is to compare the enrichment of cell markers between tissue vs urine samples. The cell markers are downloaded from the public databases [CellMarker 2.0](http://www.bio-bigdata.center/index.html) and [PanglaoDB](https://panglaodb.se/index.html).

# Preparation

```{r message=FALSE, warning=FALSE}
library(magrittr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(stringr)
library(reshape2)
library(rstatix)
library(openxlsx)
library(GSVA)
library(biomaRt)
```

```{r}
setwd("E:/study/Queens/OneDrive - Queen's University/project codes and results/9. tissue vs urine re-comparison")
```

```{r}
load("9.2. tissue vs urine re-comparison--cell markers.RData")
```

# Load data

```{r}
load("9.1. tissue vs urine re-comparison--ES and MSigDB.RData")

rm(list = ls(all.names = T)[
  !ls(all.names = T) %in% 
    c("ensembl","gene.id","sample.info","mtu.sample",
      "GSVA.expset","exp.proc.mtu","exp.proc.mtu.long",
      "exp.mtu.rep_comb","exp.mtu.rep_comb.long","list.293",
      "GSVA.anno.sam","exp.ori.sample_row",
      "exp.ori.sample_col")])
```

# CellMarker (Harbin Medical University)

## Load and curate cell marker gene sets

cell marker gene sets

```{r}
cellmarker.raw <- 
  read.xlsx("data/Harbin Med Univ CellMarker 2.0 cell markers.xlsx")

# Extract only specific columns
cellmarker.curated <- 
  cellmarker.raw %>% 
  dplyr::select(
    c(tissue_class,tissue_type,
      cancer_type,cell_type,cell_name,
      marker,Symbol,GeneID,Genetype,Genename)) %>% 
  ## Integrate tissue class and type for filtration.
  ## Also integrate tissue class, type and cell name for 
  ## identifying same cells from different organs.
  dplyr::mutate(
    class_type = paste0(tissue_class,"_",tissue_type),
    full.name = paste0(class_type,"_",cell_name))
```

Filter tissue types based on the location. The filtration results are recorded in a separate xlsx file.

```{r eval=FALSE}
# Summarize all organs and write into an xlsx file.
#
# write.xlsx(
#   cellmarker.curated[c("tissue_class","tissue_type","class_type")] %>%
#     unique %>%
#     data.frame(),
#   file = "data/organs.xlsx",
#   overwrite = T)

# Load filtration results
cellmarker.organ.filt <- 
  read.xlsx(
    xlsxFile = "data/cell marker organ filtration.xlsx") %>% 
  dplyr::filter(!is.na(keep))

# filter cell markers
cellmarker.curated %<>% 
  dplyr::filter(
    class_type %in% cellmarker.organ.filt$class_type)

cellmarker.organ.filt$class_type
```

FIlter cancer types. Include only normal and prostate cancer cells.

```{r}
cellmarker.curated %<>% 
  dplyr::filter(
    cancer_type %in% 
      c("Normal","Prostate Cancer (circulating)",
        "Prostate Cancer","Undefined",
        "Urothelial carcinoma",
        "Neuroendocrine prostate cancer",
        "Castration-resistant prostate cancer",
        "Prostate cancer"))
```

Check if our genes are in the cell marker gene sets

```{r}
gene.list.exp <- 
  data.frame(
    Gene = rownames(GSVA.expset)) %>% 
  dplyr::mutate(
    ID = gene.id$ID[
      match(Gene,gene.id$Gene)])

table(gene.list.exp$ID %in% cellmarker.curated$GeneID)
```

There are 77 genes in the marker gene sets and 290 genes not in the gene sets

Check if there are genes that do not match the symbol in the gene sets in the included genes.

```{r}
gene.list.exp[
  gene.list.exp$ID %in% cellmarker.curated$GeneID &
    !gene.list.exp$Gene %in% cellmarker.curated$Symbol,]
```

There are three genes that do not match the symbols in the gene sets.

-   ID: 5170. In the expression data: PDPK1. In the gene sets: PDK1. PDK1 is an alias of PDPK1. Keep both terms.

-   ID: 5743. In the expression data: PTGS2. In the gene sets: COX2. COX2 is an alias of PTGS2. Keep both terms.

-   ID: 23428. In the expression data: SLC7A8. In the gene sets: LAT2. LAT2 is an alias of SLC7A8. Keep both terms.

-   ID: 25840. In the expression data: TMT1A. In the gene sets: METTL7A. METTL7A is is an alias of TMT1A. Keep both terms.

Because there are multiple gene symbols matching one ID in the cell marker set, we transform gene symbols into IDs in the expression set and use IDs for GSVA.

## Organize GSVA input.

```{r}
# Gene set
GSVA.geneset.cellmarker <- 
  cellmarker.curated %>% 
  # Split the genes into a gene set list
  split(x = .$GeneID,
        f = .$full.name) %>% 
  # Remove duplicate genes
  lapply(unique)

# Expression set. Transform symbols to IDs.
rownames(GSVA.expset) <- 
  gene.id$ID[
    match(rownames(GSVA.expset),gene.id$Gene)]
```

## Perform GSVA

```{r}
overlap.gene.num <- 3

GSVA.result.cellmarker.all_sam <- 
  gsvaParam(
    exprData = GSVA.expset,
    geneSets = GSVA.geneset.cellmarker,
    kcdf = "none",
    maxDiff = T,
    absRanking = T,
    minSize = overlap.gene.num) %>%
  gsva(verbose = T) %>% 
  data.frame(check.names = F)
```

Generate a long version

```{r}
# Add a cell marker column
GSVA.result.cellmarker.all_sam %<>% 
  dplyr::mutate(
    cell.type = rownames(.))

# Generate a long version
GSVA.result.cellmarker.all_sam.long <- 
  GSVA.result.cellmarker.all_sam %>% 
  reshape2::melt(
    id.vars = "cell.type",
    variable.name = "sample.id",
    value.name = "enrichment.score") %>% 
  ## Add sample type annotations
  dplyr::mutate(
    sample.type =
      GSVA.anno.sam$`sample type`[
        match(.$sample.id,
              rownames(GSVA.anno.sam))])
```

### MTU sample results

Combine replicates by averaging each gene set so that each MTU patient has only one tissue and one urine sample.

```{r}
# Extract MTU samples
GSVA.result.cellmarker.mtu <- 
  GSVA.result.cellmarker.all_sam %>% 
  .[,colnames(.) %in% mtu.sample$SampleID.readable]

# Generate a new data frame for replicate-free enrichment matrix
for (.n.sample in 1:ncol(GSVA.result.cellmarker.mtu))
{
  ## check if the current sample is a replicate
  .sample.id <- 
    colnames(GSVA.result.cellmarker.mtu)[.n.sample]
  .patient.id <-
    sample.info$PatientID[
      match(.sample.id,
            sample.info$SampleID.readable)]
  .sample.type <-
    sample.info$SampleType[
      match(.sample.id,
            sample.info$SampleID.readable)]
  .rep.all.info <-
    sample.info %>%
    dplyr::filter(SampleType == .sample.type &
                    PatientID == .patient.id)
  
  ## Check and average the replicates.
  ## Integrate the sample into a new dataset.
  if (.n.sample == 1 & nrow(.rep.all.info) == 1)
  {
    GSVA.result.cellmarker.rep_comb <- 
      data.frame(
        GSVA.result.cellmarker.mtu[
          .rep.all.info$SampleID.readable])
  }
  else if (.n.sample == 1 & nrow(.rep.all.info) > 1)
  {
    GSVA.result.cellmarker.rep_comb <- 
      data.frame(
        GSVA.result.cellmarker.mtu[
          .rep.all.info$SampleID.readable] %>%
          apply(X = .,MARGIN = 1,FUN = mean))
  }
  else if (.n.sample > 1 & nrow(.rep.all.info) == 1)
    {
    GSVA.result.cellmarker.rep_comb %<>%
      cbind(
        GSVA.result.cellmarker.mtu[
          .rep.all.info$SampleID.readable])
    }
  else if (.n.sample > 1 & nrow(.rep.all.info) > 1)
    {
    if (.patient.id != .last.sample["patient.id"] |
        .sample.type != .last.sample["sample.type"])
      {
      GSVA.result.cellmarker.rep_comb %<>% 
        cbind(
          GSVA.result.cellmarker.mtu[
            .rep.all.info$SampleID.readable] %>%
            apply(X = .,MARGIN = 1,FUN = mean))
    }}
  
  ## Rename the new column by "patient ID + sample type"
  colnames(GSVA.result.cellmarker.rep_comb)[
    ncol(GSVA.result.cellmarker.rep_comb)] <-
    .rep.all.info$ID_type[1]
  
  ## Save the current sample type and patient ID. If the next
  ## sample is a replicate, we will skip it.
  .last.sample <- 
    c("patient.id" = .patient.id,
      "sample.type" = .sample.type)
}

# Generate a long version
GSVA.result.cellmarker.rep_comb.long <- 
  GSVA.result.cellmarker.rep_comb %>% 
  t %>% 
  data.frame(check.names = F) %>% 
  ## Add sample ID, type annotations, and patient ID
  dplyr::mutate(
    sample.id = rownames(.)) %>% 
  dplyr::mutate(
    sample.type = 
      case_when(
        str_detect(.$sample.id,"T") ~ "tissue",
        str_detect(.$sample.id,"U") ~ "urine"),
    patient.id =
      str_extract(.$sample.id,"Q-\\d+")) %>% 
  # Change to a long version
  reshape2::melt(
    id.vars = c("sample.id","sample.type","patient.id"),
    variable.name = "cell.type",
    value.name = "enrichment.score") %>% 
  # Compare the enrichment scores in tissue vs urine and 
  # add a column to record the result. For the convenience 
  # of processing, jaxtapose the enrichment scores in
  # tissue and urine first.
  reshape2::dcast(
    patient.id + cell.type ~ sample.type,
    value.var = "enrichment.score") %>% 
  # Add a result column
  dplyr::mutate(
    tissue.vs.urine = 
      if_else(tissue > urine,
              "higher in tissue",
              "higher in urine")) %>% 
  # Change the data again to the long version
  reshape2::melt(
    id.vars = c("patient.id","cell.type","tissue.vs.urine"),
    measure.vars = c("tissue","urine"),
    variable.name = "sample.type",
    value.name = "enrichment.score")

rm(.n.sample,.sample.id,.patient.id,.sample.type,
   .rep.all.info,.last.sample)
```

## Statistical tests

### Normality test

All samples

```{r}
GSVA.stat_norm.cellmarker.all_sam <- 
  GSVA.result.cellmarker.all_sam.long %>% 
  group_by(sample.type,cell.type) %>% 
  shapiro_test(vars = "enrichment.score")

table(GSVA.stat_norm.cellmarker.all_sam$p > 0.05)
```

MTU samples

```{r}
GSVA.stat_norm.cellmarker.rep_comb <- 
  GSVA.result.cellmarker.rep_comb.long %>% 
  group_by(sample.type,cell.type) %>% 
  shapiro_test(vars = "enrichment.score")

table(GSVA.stat_norm.cellmarker.rep_comb$p > 0.05)
```

There are some groups of data that are not normally distributed. Perform non-parametric tests.

### Wilcoxon test

All samples

```{r}
GSVA.stat.cellmarker.all_sam.t_vs_u <- 
  GSVA.result.cellmarker.all_sam.long %>%
  group_by(cell.type) %>% 
  rstatix::wilcox_test(
    enrichment.score ~ sample.type) %>% 
  adjust_pvalue(
    method = "BH",
    output.col = "p.adj") %>% 
  add_significance()
```

MTU samples

```{r}
GSVA.stat.cellmarker.rep_comb.t_vs_u <- 
  GSVA.result.cellmarker.rep_comb.long %>%
  group_by(cell.type) %>% 
  rstatix::wilcox_test(
    enrichment.score ~ sample.type,
    paired = T) %>% 
  adjust_pvalue(
    method = "BH",
    output.col = "p.adj") %>% 
  add_significance()
```

### Extract significant gene sets

```{r}
# statistical data
GSVA.stat.cellmarker.all_sam.t_u_diff <- 
  GSVA.stat.cellmarker.all_sam.t_vs_u %>% 
  dplyr::filter(
    p.adj < 0.05) %>% 
  ## Add and adjust y coordinates
  add_y_position() %>%
  dplyr::mutate(
    y.position = y.position + 0.1) %>% 
  ## Add x coordinates
  dplyr::mutate(
    x = seq(1,nrow(.),1),
    xmin = x - 0.2,
    xmax = x + 0.2)

GSVA.stat.cellmarker.rep_comb.t_u_diff <- 
  GSVA.stat.cellmarker.rep_comb.t_vs_u %>% 
  dplyr::filter(
    p.adj < 0.05) %>% 
  ## Add and adjust y coordinates
  add_y_position() %>% 
  dplyr::mutate(
    y.position = y.position + 0.15) %>% 
  ## Add x coordinates
  dplyr::mutate(
    x = seq(1,nrow(.),1),
    xmin = x - 0.2,
    xmax = x + 0.2)

# Long expression data
GSVA.result.cellmarker.t_u_diff.all_sam.long <- 
  GSVA.result.cellmarker.all_sam.long %>% 
  dplyr::filter(
    cell.type %in% 
      GSVA.stat.cellmarker.all_sam.t_u_diff$cell.type)

GSVA.result.cellmarker.t_u_diff.rep_comb.long <- 
  GSVA.result.cellmarker.rep_comb.long %>% 
  dplyr::filter(
    cell.type %in% 
      GSVA.stat.cellmarker.rep_comb.t_u_diff$cell.type) %>% 
  # Add x ordinates for ggplot
  dplyr::mutate(
    geom.line.x =
      case_when(
        cell.type == "Blood_Peripheral blood_Fibroblast" &
          sample.type == "tissue" ~ 0.8,
        cell.type == "Blood_Peripheral blood_Fibroblast" &
          sample.type == "urine" ~ 1.2,
        cell.type == 
          "Blood_Peripheral blood_Regulatory T (Treg) cell" &
          sample.type == "tissue" ~ 1.8,
        cell.type == 
          "Blood_Peripheral blood_Regulatory T (Treg) cell" &
          sample.type == "urine" ~ 2.2,
        cell.type == "Prostate_Prostate_Basal cell" &
          sample.type == "tissue" ~ 2.8,
        cell.type == "Prostate_Prostate_Basal cell" &
          sample.type == "urine" ~ 3.2))
```

```{r}
GSVA.stat.cellmarker.all_sam.t_u_diff$cell.type
```

```{r}
GSVA.stat.cellmarker.rep_comb.t_u_diff$cell.type
```

There are three gene sets with significant differences between tissue and urine samples.

## Visualization

All samples

```{r fig.height=6, fig.width=8}
ggplot(
  data = GSVA.result.cellmarker.t_u_diff.all_sam.long,
  mapping = aes(
    x = cell.type,y = enrichment.score,
    fill = sample.type)) +
  stat_summary(
    fun.min = function(x){quantile(x)["75%"]+1.5*sd(x)},
    fun.max = function(x){quantile(x)["25%"]-1.5*sd(x)},
    geom = "errorbar",
    position = position_dodge(width = 0.8),
    width = 0.4,color = "black",linewidth = 0.5) +
  geom_boxplot(
    position = position_dodge(0.8),
    coef = 0,
    outlier.shape = NA) +
  scale_fill_manual(
    labels = c(
      paste0("tissue\n(n=",
             GSVA.stat.cellmarker.all_sam.t_vs_u$n1[1],")"),
      paste0("urine\n(n=",
             GSVA.stat.cellmarker.all_sam.t_vs_u$n2[1],")")),
    values = c("darkmagenta","yellow3")) +
  xlab(NULL) +
  ylab(NULL) +
  scale_x_discrete(
    labels = c("vascular\nendothelial cell",
               "blood\nfibroblast",
               "blood\nTreg cell",
               "prostate\nbasal cell")) +
  ggtitle("enrichment scores of cell marker genes\nin all samples\ntissue vs urine") +
  stat_pvalue_manual(
    GSVA.stat.cellmarker.all_sam.t_u_diff,
    label = "p.adj.signif",
    tip.length = 0.01,label.size = 7) +
  theme(
    axis.line = element_line(),
    panel.background = element_blank(),
    plot.title = element_text(
      size = 20,face = "bold",hjust = 0),
    axis.text.x = element_text(
      size = 20,hjust = 0.5),
    axis.text.y = element_text(
      size = 20),
    legend.title = element_blank(),,
    legend.key = element_rect(
      fill = "white",color = "white"),
    legend.key.height = unit(1,"cm"),
    legend.key.width = unit(0.5,"cm"),
    legend.text = element_text(size = 20))
```

MTU samples

```{r}
ggplot(
  data = GSVA.result.cellmarker.t_u_diff.rep_comb.long,
  mapping = aes(
    x = cell.type,y = enrichment.score,
    fill = sample.type)) +
  stat_summary(
    fun.min = function(x){quantile(x)["75%"]+1.5*sd(x)},
    fun.max = function(x){quantile(x)["25%"]-1.5*sd(x)},
    geom = "errorbar",
    position = position_dodge(width = 0.8),
    width = 0.4,color = "black",linewidth = 0.5) +
  geom_boxplot(
    position = position_dodge(0.8),
    coef = 0,
    outlier.shape = NA) +
  scale_fill_manual(
    labels = c(
      paste0("tissue\n(n=",
             GSVA.stat.cellmarker.rep_comb.t_vs_u$n1[1],")"),
      paste0("urine\n(n=",
             GSVA.stat.cellmarker.rep_comb.t_vs_u$n2[1],")")),
    values = c("darkmagenta","yellow3")) +
  geom_point(
    position = position_dodge(0.8)) +
  geom_line(
    data = 
      GSVA.result.cellmarker.t_u_diff.rep_comb.long %>% 
      dplyr::filter(
        cell.type == 
          "Blood_Peripheral blood_Fibroblast"),
    mapping = aes(
      x = geom.line.x,y = enrichment.score,
      group = patient.id,color = tissue.vs.urine)) +
  geom_line(
    data = 
      GSVA.result.cellmarker.t_u_diff.rep_comb.long %>% 
      dplyr::filter(
        cell.type == 
          "Blood_Peripheral blood_Regulatory T (Treg) cell"),
    mapping = aes(
      x = geom.line.x,y = enrichment.score,
      group = patient.id,color = tissue.vs.urine)) +
  geom_line(
    data = 
      GSVA.result.cellmarker.t_u_diff.rep_comb.long %>% 
      dplyr::filter(
        cell.type == 
          "Prostate_Prostate_Basal cell"),
    mapping = aes(
      x = geom.line.x,y = enrichment.score,
      group = patient.id,color = tissue.vs.urine)) +
  scale_color_manual(
    labels = c("higher in tissue",
               "higher in urine"),
    values = c("dodgerblue","sienna")) +
  xlab(NULL) +
  ylab(NULL) +
  ggtitle("enrichment scores of cell marker genes\nin MTU samples\ntissue vs urine") +
  scale_x_discrete(
    labels = c("blood\nfibroblast",
               "blood\nTreg cell",
               "prostate\nbasal cell")) +
  stat_pvalue_manual(
    GSVA.stat.cellmarker.rep_comb.t_u_diff,
    label = "p.adj.signif",
    tip.length = 0.01,label.size = 7) +
  theme(
    axis.line = element_line(),
    panel.background = element_blank(),
    plot.title = element_text(
      size = 20,face = "bold",hjust = 0),
    axis.text.x = element_text(
      size = 20,hjust = 0.5),
    axis.text.y = element_text(
      size = 20),
    legend.title = element_blank(),,
    legend.key = element_rect(
      fill = "white",color = "white"),
    legend.key.height = unit(1,"cm"),
    legend.key.width = unit(0.5,"cm"),
    legend.text = element_text(size = 20))
```

## Summary

In either all samples or MTU samples, three types of cell markers were significantly enriched in urine compared to tissue:
